---
title: 'BIOS 333 — Week 1 In‑Class: Getting Comfortable with ggplot2'
author: "Lizbeth Macias"
date: "`r format(Sys.Date(), '01 29, 26')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
    theme: readable
  pdf_document:
    toc: true
    toc_depth: '2'
---

## Goals for today (read this first)

By the end of class you should be able to:

- Build a plot from scratch using `ggplot()` + layers (geoms).
- Map variables to aesthetics (`x`, `y`, `color`, `size`, `shape`, `alpha`).
- Debug common problems by *breaking* a plot on purpose and then fixing it.
- Use themes and labels to improve readability.
- Write short, clear notes that interpret what your code did and what your plot shows.

**Important:** This is an in‑class assignment. You must fill in the **TODO** sections, knit to HTML, and submit the HTML to Blackboard.

## Required reading (for context, not a quiz)

You should have read (or skimmed) these before class:

- Color basics: https://clauswilke.com/dataviz/color-basics.html  
- Aesthetic mapping: https://clauswilke.com/dataviz/aesthetic-mapping.html  
- Coordinate systems & axes: https://clauswilke.com/dataviz/coordinate-systems-axes.html  
- Healy (socviz) make a plot: https://socviz.co/makeplot.html#makeplot  

---

## Setup

### 0) Your notes (required)

In 3–5 bullet points, write what you think ggplot2 is doing when you make a plot.

**TODO (write your notes below):**

- You tell ggplot which variables go where by mapping them to aesthetics like color, size, etc.
- Themes and labels make the plot easier to read, without changing the data itself.
- ggplot2 builds a plot in layers: you start with the data, then add pieces (like points or lines) one layer at a time.

### 1) Packages

We will use `ggplot2` and a built‑in dataset (`mtcars`) so there is **no data download** needed.

```{r setup, message=FALSE, warning=FALSE}
# If you are running this on a computer that does NOT have ggplot2 installed yet,
# uncomment the next line once:
# install.packages("ggplot2")

library(ggplot2)

# Optional: for nicer data manipulation later (not required today)
# install.packages("dplyr")
# library(dplyr)
```

### 2) Load / inspect the dataset

`mtcars` is built into R. Let's look at it.

```{r}
# TODO: Run these lines and interpret what you see.

# View the first few rows
# head(mtcars)

# Look at column names
# names(mtcars)

# Get a quick summary
# summary(mtcars)
```

**Interpretation (2–4 sentences):**  
**TODO:** What does one row represent? What are 2 variables you think might relate to fuel efficiency (`mpg`)?

- One row represents one car are its characteristics measured since each column shows a unique feature like mpg or weight. 
- 2 variables I think might relate to fuel efficiency are weight (wt) since the heavier a vehicle is, the more fuel/gas it intakes, and size of the car since the larger the car, more fuel it needs.
---

## Part A — The smallest possible ggplot

### 3) Start with a blank ggplot object


```{r}
# TODO: Create a ggplot object that uses mtcars as the data.
# Hint: ggplot(data = mtcars)

# NOTE: removed hashtags in both p to run function properly. R ignores action if hashtag is in front of it.
p <- ggplot(data = mtcars)
p
```

**Interpretation (1–2 sentences):**  
**TODO:** Why does this not show a plot yet?

- This plot does not show a plot yet since no data has been given or instructed even though the command ran with no errors. We need to specify it.

### 4) Add points (a first geom)

```{r}
# TODO: Add a scatterplot layer with geom_point()
# Requirements:
# - x axis: wt
# - y axis: mpg
# Note: removed hashtags on line ggplot and geom_point

ggplot(data = mtcars, aes(x = wt, y = mpg)) +
geom_point()
```

**Interpretation (2–4 sentences):**  
**TODO:** Describe the relationship between weight (`wt`) and miles per gallon (`mpg`). Is it positive or negative? Strong or weak?

- The relationship between wt and mpg is negative because as a car’s weight increases, fuel efficiency tends to decrease when observing the graph. Additionally, the points show a clear downward trend, showing a strong trend. 
---

## Part B — Aesthetics: map vs set
3 Part B
### 5) Map a variable to color
Note color point by number of cyinders

```{r}
# TODO: Color points by number of cylinders (cyl)
# Hint: put color = cyl inside aes(...)
# Note: removed hashtages in ggplot and geom so that R doesn't ignore command.
 ggplot(mtcars, aes(x = wt, y = mpg, color = cyl)) +
   geom_point()
#Note: make all points one color
```

**Interpretation (2–4 sentences):**  
**TODO:** What does "mapping" mean here? What does the legend represent?
- mapping means assigning a variable from the data (cyl) to create a an aesthetic for (color). Each point on the graph colors are determined by how many cylinders a particular car has. The legend shows which color goes to cylinder group it belongs in. 
 

### 6) Set a constant color (NOT mapped)

```{r}
# TODO: Make points blue (or any single color you like) BUT do not create a legend.
# Hint: put color = "blue" outside aes(...)
# Note: removed hashtags to make all plots all blue
 ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(color = "blue")
```

**Interpretation (1–2 sentences):**  
**TODO:** Explain why the previous plot had a legend but this one does not.
- The previous plot had a legend because color was mapped to a variable (cyl) inside aes(), so ggplot needed a legend to explain the different colors. 
- In this plot, the color is set to a single value being ("blue") that is outside aes(), so it does not represent any data. 
### 7) Try other aesthetics

Choose ONE:
- Map `size` to `hp`, OR
- Map `shape` to `gear`, OR
- Map `alpha` to `disp`

```{r}
# TODO: Make points green (or any single color you like) BUT do not create a legend.
# Hint: put color = "green" outside aes(...)

 ggplot(mtcars, aes(x = wt, y = mpg, color = cyl, shape = factor ( gear ))) + geom_point()
```

**Interpretation (2–4 sentences):**  
**TODO:** Was your chosen aesthetic helpful or confusing? Why?
- My chosen aesthetic was confusing since changing the shape to gear makes it visually difficult to understand what the data is interpreting even though a legend is present.
---

## Part C — Break it on purpose, then fix it

### 8) "Object not found" in ggplot2

Run the code below **as is**. You should get an error.

```{r}
# DO NOT FIX THIS LINE YET — run it first and read the error.

#Note: added hashtags since it should give you an error but idn't want to have issues when knitting

# ggplot(mtcars, aes(x = weight, y = mpg)) +
#  geom_point()
```

**TODO:** Copy/paste the exact error message here:

Error in `geom_point()`:
! Problem while computing aesthetics.
ℹ Error occurred in the 1st layer.
Caused by error:
! object 'weight' not found
Run `rlang::last_trace()` to see where the error occurred.
> 

Now fix it.

```{r}
# TODO: Fix the plot by using the correct column name.
# Hint: check names(mtcars)
# Note: changed weight to x= wt because it doesn't know weight
# Your fixed code here
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point()
```

**Interpretation (2–4 sentences):**  
**TODO:** What did the error *mean* in plain English? How did you diagnose it?

- The "error" means that I asked R to use a variable/function, that it cannot locate in its current memory, in this case it couldn't locate (wt) since it was written as "weight".
---

## Part D — Layers: add a trend line, then modify it

### 9) Add a smooth trend
Note: removed hashtags in lines 218 to 220
```{r}
# TODO: Add geom_smooth() to your scatterplot of wt vs mpg
# Hint: add a new layer with +
# Note: a solid line and a surrounding grey area usually represent a regression line and its confidence interval produced by geom_smooth() function. 
 ggplot(mtcars, aes(wt, mpg)) +
   geom_point() +
   geom_smooth()
```

### 10) Make the smooth line a *linear model*

```{r}
# TODO: Modify geom_smooth so it fits a straight line.
# Hint: geom_smooth(method = "lm", se = FALSE)
# Note: what does se: false do? turns confidence intervals off
# Note: added new code for smooth linear line
# Your code here
ggplot(mtcars, aes (x = wt, y= mpg)) +
   geom_point() +
   geom_smooth(method ="lm", se = FALSE)
```

**Interpretation (2–4 sentences):**  
**TODO:** What does `se = FALSE` change? Why might you turn it off?
- 'se = FALSE' removes the shaded confidence interval ribbon around the regression line to make the plot look cleaner visually.
---

## Part E — Facets: small multiples

### 11) Compare groups with facets

```{r}
# TODO: Facet by cylinders (cyl)
# Hint: + facet_wrap(~ cyl)
# Note: edit all this
# Your code here
ggplot(mtcars, aes(wt, mpg, color = factor (cyl), shape = factor (gear))) + geom_point() +
   geom_smooth(method ="lm", se = FALSE) 
   facet_wrap(~cyl)
```

**Interpretation (2–4 sentences):**  
**TODO:** In what way is faceting different from coloring? Which is clearer here?
- faceting allows to plot in grids to show a different subset of data in a specific category.
- coloring is more clearer when showing differences and direct comparisons while faceting prevents overlap of data points.
---

## Part F — Labels, themes, and coordinate systems

### 12) Make it readable (labels + theme)

```{r}
# TODO: Improve your plot.
# Requirements:
# - Add a title
# - Label x and y axes with words (not just wt/mpg)
# - Use theme_minimal() OR theme_classic()

# Your code here
   ggplot(mtcars, aes(wt, mpg, color = factor(cyl), shape = factor(gear))) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ cyl) +
  labs(
    title = "Fuel Efficiency vs Car Weight by Cylinder",
    x = "Car Weight (1000 lbs)",
    y = "Miles per Gallon (mpg)"
  ) +
  theme_minimal()
   
```

**Interpretation (2–4 sentences):**  
**TODO:** What design choices did you make and why?
- Design choices I made to this graph is include a title on the top, title on both (x and y) axis. 
- This helped make the graph more visually appealing but also include the necessary information it needed to understand what the plot is interpreting.
### 13) Coordinate systems: zoom vs change scale

**A. Zoom (keep the data, change what you *see*)**

```{r}
# TODO: Add coord_cartesian(ylim = c(10, 35)) (or similar)
# Use your favorite plot from above.
# Note: addded coord_cartesian as extended function in chunk. Not really sure what it did.
# Your code here
ggplot(mtcars, aes(wt, mpg, color = factor(cyl), shape = factor(gear))) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ cyl) +
  labs(
    title = "Fuel Efficiency vs Car Weight by Cylinder",
    x = "Car Weight (1000 lbs)",
    y = "Miles per Gallon (mpg)"
  ) +
  coord_cartesian(ylim = c(10, 35)) +
  theme_minimal()
```

**B. Filter (remove data before plotting)**

```{r}
# TODO: Create a filtered dataset that keeps only cars with mpg <= 35 (or similar),
# then plot it. You may use base R or dplyr.

# Option 1 (base R):
# mt_small <- mtcars[mtcars$mpg <= 35, ]

# Option 2 (dplyr):
# mt_small <- mtcars %>% filter(mpg <= 35)
# Note: option 1: 32 observations empty_small of 11 variables
# Note: add $
# Your code here
# Note: did option 1

ggplot(mtcars, aes(wt, mpg, color = factor(cyl), shape = factor(gear))) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ cyl) +
  labs(
    title = "Fuel Efficiency vs Car Weight by Cylinder",
    x = "Car Weight (1000 lbs)",
    y = "Miles per Gallon (mpg)"
  ) +
  coord_cartesian(ylim = c(10, 35)) +
  theme_minimal()

  mt_small <- mtcars[mtcars$mpg <= 35, ]

ggplot(mt_small, aes(wt, mpg)) +
  geom_point()

```

**Interpretation (3–5 sentences):**  
**TODO:** Explain the difference between *zooming* with `coord_cartesian()` and *filtering* the data first. Why does this matter ethically?
- "Zooming" with coord_cartesian() restricts how we view the window of a plot while keeping the entire dataset intact
- It acts as a magnifier. Examples could be trends, smoothers, etc. - "filtering" data first removes data points outside the specified range before plotting
- This can change the statistical summaries and visual interpretations of the graphs created. This matters ethically because filtering unaware can lead to the removal of outliers or contextual information.

---

## Part G — Final challenge (choose one)

Pick ONE challenge below and complete it. Your goal is to show you can build a plot, break it, and rebuild it.

### Option 1: Boxplot + jitter
Note: add boxplot and jitter piece
Note: geombox + jitter code
Note: x=cylinder
y=mpg
Note: not doing geom or facet
Make a boxplot of `mpg` by `cyl`, then overlay jittered points.
# Note: included geom_boxplot and geom_jitter
```{r}
# TODO: Your code here
ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
  geom_boxplot() +
  geom_jitter(width = 0.3)
```

**Interpretation (2–4 sentences):**  
**TODO:** What does the boxplot summarize that the points do not (or vice versa)?
- Boxplot summarizes and shows points in the median, quartiles and interquartile range (IQR).
- In doing so, it provides immediate visual cues for overall spread and skewness, which can be difficult to tell without it.

---

## Turn‑in checklist

Before you submit:

- [ ] You knitted to **HTML** with no errors.
- [ ] Every **TODO** section has either code or prose.
- [ ] Your plots appear in the knitted HTML.
- [ ] Your interpretations are in your own words.

## Session info (required)

```{r}
sessionInfo()
```
